Interacting with Modal Logics \\ in the Coq Proof Assistant

### 2014-12-11 Papers/InProgress/ModalLogicCoq/ModalCoq.tex
Improve language of sections 6 and 7

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology. G\"{o}del's proof \cite{Goedel1970,GoedelNotes} is a modern
culmination of this tradition, following particularly the footsteps of
Leibniz. Various slightly different versions of axioms and definitions
have been considered by G\"{o}del and by several philosophers who
commented on his proof (cf.
\cite{sobel2004logic,AndersonGettings,Fitting,Adams,ContemporaryBibliography}).
The formalization shown in this Section aims at being as similar as
possible to Dana Scott's version of the proof \cite{ScottNotes}. The
formulation and numbering of axioms, definitions and theorems is the
same as in Scott's notes. Even the \Coq proof scripts follow precisely
all the steps in Scott's notes. Scott's assertions are emphasized
below with comments. In contrast to the formalization in \Isabelle
\cite{AFP}, where automation via \texttt{Metis}
\cite{Hurd03first-orderproof} and \texttt{Sledgehammer}
\cite{Sledgehammer} using tools such as \texttt{Nitpick}
\cite{Nitpick}, \texttt{LEO-II} \cite{LEO-II} and \texttt{Satallax}
\cite{Satallax} has been successfully employed, the formalization in
\Coq used no automation. This was a deliberate choice, mainly because
it allowed a qualitative evaluation of the convenience of the
embedding approach for \emph{interactive} theorem proving. Moreover,
in order to formalize precisely Scott's version and not another
automatically found version, automation would have to be heavily
limited anyway. Furthermore, the deliberate preference for simple
tactics (mostly \emph{intro}, \emph{apply} and the modal tactics
described in Section \ref{sec:Tactics}) results in proof scripts that
closely correspond to natural deduction proofs. This hopefully makes
the formalization more accessible to those who are not experts in
\Coq's tactics but are nevertheless interested in G\"odel's proof.

@@

\noindent 
In Scott's notes, classicality occurs in uses of the
principle of proof by contradiction. Therefore, in order to clearly
indicate where classical logic is needed in the proof scripts, a
simple tactic that simulates proof by contradiction can be created:

@@

\noindent
Five axioms are stated by Gödel to characterize positivity. 
The first three axioms are needed to establish the first theorem and 
its corollary stating the possibility of God's existence, 
with God defined as a being possessing all positive properties.

@@

\noindent 
The second part of the proof consists in showing that if
God's existence is possible then it must be necessary
(\texttt{lemma2}), for which the controversial \textbf{S5} principle
\texttt{dia\_box\_to\_box} is used.

@@

The successful formalization of Scott's version of  G\"odel's
ontological argument indicates that the \emph{embedding} of
higher-order modal
logics into higher-order logics via the \emph{possible world
semantics} is a viable approach for fully interactive theorem proving
within (even non-trivial) modal logics. Our lightweight implementation of the embedding
(described in Sections \ref{sec:Embedding} and \ref{sec:Tactics})
takes special care to hide the underlying possible world machinery
from the user. An inspection of the proof scripts in Section
\ref{sec:proof} shows that this goal has been achieved. The user does
not have to explicitly bother about worlds and their mutual
reachability; the provided tactics for modalities do the job for
him/her. Moreover, for subgoals that do not involve modalities, the
user has all the usual interactive tactics at his/her disposal.

>>>>

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology. G\"{o}del's proof \cite{Goedel1970,GoedelNotes} is a modern
culmination of this tradition, following particularly the footsteps of
Leibniz. Various slightly different versions of axioms and definitions
have been considered by G\"{o}del and by several philosophers who
commented on his proof (cf.
\cite{sobel2004logic,AndersonGettings,Fitting,Adams,ContemporaryBibliography}).
The formalization shown in this Section aims at being as similar as
possible to Dana Scott's version of the proof \cite{ScottNotes}. The
formulation and numbering of axioms, definitions and theorems is the
same as in Scott's notes. Even the \Coq proof scripts follow precisely
all the steps in Scott's notes. Scott's assertions are emphasized
below with comments. In contrast to the formalization in \Isabelle
\cite{AFP}, where automation via \texttt{Metis}
\cite{Hurd03first-orderproof} and \texttt{Sledgehammer}
\cite{Sledgehammer} using tools such as \texttt{Nitpick}
\cite{Nitpick}, \texttt{LEO-II} \cite{LEO-II} and \texttt{Satallax}
\cite{Satallax} has been successfully employed, the formalization in
\Coq used no automation. This was a deliberate choice, mainly because
it allowed a qualitative evaluation of the convenience of the
embedding approach for \emph{interactive} theorem proving. Moreover,
in order to formalize exactly Scott's version and not some arbitrary 
version found automatically, automation would have to be heavily
limited anyway. Furthermore, the deliberate preference for simple
tactics (mostly \emph{intro}, \emph{apply} and the modal tactics
described in Section \ref{sec:Tactics}) results in proof scripts that
closely correspond to common natural deduction proofs. This hopefully makes
the formalization more accessible to those who are not experts in
\Coq's tactics but are nevertheless interested in G\"odel's proof.

@@

\noindent 
In Scott's notes, classicality occurs in uses of the
principle of proof by contradiction. In order to clearly
indicate where classical logic is needed in the proof scripts, a
simple tactic that simulates proof by contradiction was created:

@@

\noindent
God is defined as a being possessing all positive properties, and
five axioms are stated to characterize positivity. 
The first part of the proof culminates in \texttt{corollary1} and establishes
that God's existence is possible.

@@

\noindent 
The second part of the proof consists in showing that if
God's existence is possible then it must be necessary
(\texttt{lemma2}). The controversial \textbf{S5} principle
\texttt{dia\_box\_to\_box} is used.

@@

The successful formalization of Scott's version of  G\"odel's
ontological argument indicates that the \emph{embedding} of
higher-order modal
logics into higher-order logics via the \emph{possible world
semantics} is a viable approach for fully interactive theorem proving
within modal logics. Our lightweight implementation of the embedding
(described in Sections \ref{sec:Embedding} and \ref{sec:Tactics})
takes special care to hide the underlying possible world machinery
from the user. An inspection of the proof scripts in Section
\ref{sec:proof} shows that this goal has been achieved. The user does
not have to explicitly bother about worlds and their mutual
reachability; the provided tactics for modalities do the job for
him/her. Moreover, for subgoals that do not involve modalities, the
user has all the usual interactive tactics at his/her disposal.
